---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  comment = "#>",
  fig.path = "README-"
)
```

# smoothr

The goal of `smoothr` is to offer a variety of methods for smoothing spatial features (i.e. polygons and lines). This package offers support for both `sp` and `sf` spatial objects, and the following smoothing methods:

- **Bézier splines**: smoothing using Bézier spline interpolation via the `spline()` function. This method interpolates between existing vertices and the resulting smoothed feature will pass through the vertices of the input feature.

## Installation

You can install smoothr from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("mstrimas/smoothr")
```

## Example

Two example feature sets are included in this package. `jagged_polygons` contains 9 polygons with sharp edges for smoothing, some have holes and some are multipart polygons. These can be smoothed with:

```{r smooth-polygons}
library(sf)
library(smoothr)
par(mar = c(0, 0, 0, 0), oma = c(0, 0, 2, 0), mfrow = c(3, 3))
for (i in 1:nrow(jagged_polygons)) {
  p <- jagged_polygons[i, ]
  smoothed <- smooth(p, method = "spline")
  plot(st_geometry(smoothed), col = NA, border = NA)
  plot(st_geometry(p), col = "grey20", border = NA, add = TRUE)
  plot(st_geometry(smoothed), col = NA, border = "red", lwd = 2, add = TRUE)
  title("Smoothed Polygons (Bézier Spline)", cex.main = 2, outer = TRUE)
}
```

`jagged_polygons` contains 9 lines with sharp edges for smoothing, some are closed loops requiring special treatment of the endpoints and some are multipart lines. These can be smoothed with:

```{r smooth-lines}
par(mar = c(0, 0, 0, 0), oma = c(0, 0, 2, 0), mfrow = c(3, 3))
for (i in 1:nrow(jagged_lines)) {
  l <- jagged_lines[i, ]
  smoothed <- smooth(l, method = "spline")
  plot(st_geometry(smoothed), col = NA)
  plot(st_geometry(l), col = "grey20", lwd = 2, add = TRUE)
  plot(st_geometry(smoothed), col = "red", lwd = 2, add = TRUE)
  title("Smoothed Lines (Bézier Spline)", cex.main = 2, outer = TRUE)
}
```
